Ecological Metadata Language interface for R: facilitating synthesis and integration of heterogenous data

### 2014-04-22 inst/doc/pubs/one-simple-way.Rmd

## 3. Provide unprocsessed data

@@

@@

The detailed column and unit metadata serve the pupose described in this section of White _et al_.   
Machine-readable metadata,

>>>>

## 3. Provide unprocessed data

@@

Data tables are easiest to manipulate and combine when users consistently
enter variables as columns and observations as rows (sometimes called "long" format).  Partly because this format
may require duplicated entries, such as repeating the value of the "site" when
multiple observations have been made on the same site, many researchers are 
still tempted to present data in "wide" format, with observations for a single 
site added as additional rows.  While potentially convenient for field entry,
the long format is more desirable for most future analysis or reuse.

The EML package addresses this by both working in the R an environment where
it is easy and routine to transform between "long" and "wide" formats, while
also giving the user the flexibility to simply document the data in whatever
format they feel works best for their analysis.

@@

As White _et al._ discuss, providing the kind of detailed column and unit
metadata we have already discussed, as well as the standard formats addressed
above, already make data easy to combine.

The EML package lets the user go beyond this in being able to automate
the combining of datasets based on the metadata alone. The key to this process
is to replace a typical programmatic script for subsetting or combining data sets 
which is based on implicit knowledge of the data structure (e.g. "select value from column 5 whenever column 2 has value = AAA") into one that is based only 
on the metadata description: (e.g. "select the species_population_counts from the Sacramento_River site").  The latter has the advantage of not only 
being easier to interpret, but can also be independent of how the data is structured.

This is already possible with R, but requires explicit use of the column name.  
As the column name may already have a specific value given to it by the researcher
that serves some other purpose, it would be against the objectives of EML to insist that it have a specific name (e.g. all columns containing species names must be called "spp", which among other issues would break some code requiring that
the column instead be called "species".)  To avoid renaming the columns, we simply
need to use a consistent description of the column in the metadata instead.
